Equalizer\\Programming and User Guide

### 2009-11-18 Developer/ProgrammingGuide/paper.tex
FSAA-related documentation update.
[ ] May break build
[ ] May break existing applications
[ ] Bugfix
[ ] New Feature
[ ] Cleanup
[ ] Optimization
[x] Documentation

@@

\subsubsection{Decomposition - Attributes}
Compounds have attributes which configure the decomposition of the
destination channel's rendering, which is defined by the viewport,
frustum and database. A \textsf{viewport} decomposes the destination
channel and frustum in screen space. A \textsf{range} tells the
application to render a part of its database, and an \textsf{eye}
rendering pass can selectively render different stereo passes. A
\textsf{pixel} parameter adjusts the frustum so that the source channel
renders an even subset of the parent's pixels. Setting one or multiple
attributes causes the parent's view to be decomposed
accordingly. Attributes are cumulative, that is, intermediate compound
attributes affect and therefore decompose the rendering of all their
children.

@@

void FrameData::serialize( eq::net::DataOStream& os, const uint64_t dirtyBits )
{
    eq::Object::serialize( os, dirtyBits );
    if( dirtyBits & DIRTY_CAMERA )
        os << _translation << _rotation << _modelRotation;
    if( dirtyBits & DIRTY_FLAGS )
        os << _modelID << _renderMode << _colorMode << _ortho << _statistics
           << _help << _wireframe << _pilotMode;
    if( dirtyBits & DIRTY_VIEW )
        os << _currentViewID;
    if( dirtyBits & DIRTY_MESSAGE )
        os << _message;
}

void FrameData::deserialize( eq::net::DataIStream& is,
                             const uint64_t dirtyBits )
{
    eq::Object::deserialize( is, dirtyBits );
    if( dirtyBits & DIRTY_CAMERA )
        is >> _translation >> _rotation >> _modelRotation;
    if( dirtyBits & DIRTY_FLAGS )
        is >> _modelID >> _renderMode >> _colorMode >> _ortho >> _statistics
           >> _help >> _wireframe >> _pilotMode;
    if( dirtyBits & DIRTY_VIEW )
        is >> _currentViewID;
    if( dirtyBits & DIRTY_MESSAGE )
        is >> _message;
}
\end{lstlisting}}%>>

@@

{\footnotesize\begin{lstlisting}
uint32_t Config::startFrame()
{
    // update head position
    if( _tracker.isRunning() )
    {
        _tracker.update();
        const vmml::Matrix4f& headMatrix = _tracker.getMatrix();
        setHeadMatrix( headMatrix );
    }

    // update database
    _frameData.data.rotation.preRotateX( -0.001f * _spinX );
    _frameData.data.rotation.preRotateY( -0.001f * _spinY );
    const uint32_t version = _frameData.commit();

    return eq::Config::startFrame( version );
}
\end{lstlisting}}

@@

\begin{description}
\item[Buffer] The OpenGL read and draw buffer as well as color mask.
  These parameters are influenced by the current eye pass, eye
  separation and anaglyphic stereo settings.
\item[Viewport] The two-dimensional pixel viewport restricting the
  rendering area within the channel. For correct operations, both
  \textsf{glViewport} and \textsf{glScissor} have to be used. The pixel
  viewport is influenced by the destination channel's viewport
  definition and compound viewports set for sort-first/2D decompositions.
\item[Frustum] The same frustum parameters as defined by
  \textsf{glFrustum}. Typically the frustum used to set up the OpenGL
  projection matrix. The frustum is influenced by the destination
  channel's view definition, compound viewports, head matrix and the
  current eye pass.
\item[Head Transformation] A transformation matrix positioning the
  frustum. This is typically an identity matrix and is used for off-axis
  frusta in immersive rendering. It is normally used to set up the
  `view' part of the modelview matrix, before static light sources are
  defined.
\item[Range] A one-dimensional range with the interval [0..1]. This
  parameter is optional and should be used by the application to render
  only the appropriate subset of its data. It is influenced by the
  compound range attribute.
\end{description}

@@

{\footnotesize\begin{lstlisting}
    const Model* model = _getModel();
    if( model )
    {
        _drawModel( model );
    }
    else
    {
        glNormal3f( 0.f, -1.f, 0.f );
        glBegin( GL_TRIANGLE_STRIP );
        glVertex3f(  .25f, 0.f,  .25f );
        glVertex3f(  .25f, 0.f, -.25f );
        glVertex3f( -.25f, 0.f,  .25f );
        glVertex3f( -.25f, 0.f, -.25f );
        glEnd();
    }
}
\end{lstlisting}}

@@

Like most applications, eqPly can use the default implementation of the
\textsf{frameReadback} and \textsf{frameAssemble} task methods. Only
\textsf{frameReadback} is overwritten for an optimization. Since eqPly
does not need the alpha channel on the destination view, the output
frames are flagged to ignore it, which allows the compressor to drop
25\% of the data during image transfer:

@@

@@

@@

The viewport restricts the rendering to the area relative to the parent
compound. The range restricts the database range, relative to the
parent. The pixel setting selects the pixel decomposition kernel,
relative to the parent. The eye pass select the mono, left or right eye
passes rendered by the compound. The zoom scales the parent pixel
viewport resolution. The DPlex period defines that $\frac{1}{period}$
frames are rendered, and the phase defines when in the period the
rendering starts. All these attributes are inherited by the children of
a compound. Viewport, range, pixel and period parameters are cumulative.

@@

    viewport [ viewport ]            # wrt parent compound, sort-first
    range    [ float float ]         # DB-range for sort-last
    pixel    [ int int ]             # pixel decomposition (step size)
    eye      [ CYCLOP LEFT RIGHT ]   # monoscopic or stereo view
    zoom     [ float float ]         # up/downscale of parent pvp
    period   int                     # DPlex period, every nth frame rendered
    phase    int                     # DPlex phase, when to start rendering

>>>>

\subsection{Subpixel Compounds}

Subpixel decomposes the rendering into different jittered frusta so 
that the recomposition anti-aliases the final rendering.
The Subpixel compound is also compatible with applications doing their own software 
anti-aliasing.

As for the DB compound, the subpixel recomposition has linear increasing 
IO requirements for the pixel transfer with the difference that only 
color information has to be transmitted.

Subpixel compounds in Equalizer are configured using the \textsf{subpixel}
parameter, using the values \textsf{[ index size ]}. The \textsf{index} 
parameter corresponds to the current ressource's index and the \textsf{size} 
is the total number of ressources needed.

The \textsf{index} has to be smaller than the ressources \textsf{size}. 
\fig{fSubpixel} illustrates a four-way subpixel decomposition.

The destination channel can also be used as a source channel. 
As for the pixel compound, it has to use an output and corresponding 
input frame.

\begin{figure}[ht!]\center
  \includegraphics[width=.9\textwidth]{images/subpixel.pdf}
  {\caption{\label{fSubpixel}A four-way Subpixel compound}}
\end{figure}

@@

\subsubsection{Decomposition - Attributes}
Compounds have attributes which configure the decomposition of the
destination channel's rendering, which is defined by the viewport,
frustum and database. A \textsf{viewport} decomposes the destination
channel and frustum in screen space. A \textsf{range} tells the
application to render a part of its database, and an \textsf{eye}
rendering pass can selectively render different stereo passes. A
\textsf{pixel} parameter adjusts the frustum so that the source channel
renders an even subset of the parent's pixels. A \textsf{subpixel} parameter
tells the source channels to render different jittered frustum to perform
anti-aliasing. Setting one or multiple attributes causes the parent's view
to be decomposed accordingly. Attributes are cumulative, that is,
intermediate compound attributes affect and therefore decompose the
rendering of all their children.

@@

void FrameData::serialize( eq::net::DataOStream& os, const uint64_t dirtyBits )
{
    eq::Object::serialize( os, dirtyBits );
    if( dirtyBits & DIRTY_CAMERA )
        os << _translation << _rotation << _modelRotation;
    if( dirtyBits & DIRTY_FLAGS )
        os << _modelID << _renderMode << _colorMode << _ortho << _statistics
           << _help << _wireframe << _pilotMode << _idleMode;
    if( dirtyBits & DIRTY_VIEW )
        os << _currentViewID;
    if( dirtyBits & DIRTY_MESSAGE )
        os << _message;
}

void FrameData::deserialize( eq::net::DataIStream& is,
                             const uint64_t dirtyBits )
{
    eq::Object::deserialize( is, dirtyBits );
    if( dirtyBits & DIRTY_CAMERA )
        is >> _translation >> _rotation >> _modelRotation;
    if( dirtyBits & DIRTY_FLAGS )
        is >> _modelID >> _renderMode >> _colorMode >> _ortho >> _statistics
           >> _help >> _wireframe >> _pilotMode >> _idleMode;
    if( dirtyBits & DIRTY_VIEW )
        is >> _currentViewID;
    if( dirtyBits & DIRTY_MESSAGE )
        is >> _message;
}
\end{lstlisting}}%>>

@@

{\footnotesize\begin{lstlisting}
uint32_t Config::startFrame()
{
    // update head position
    if( _tracker.isRunning() )
    {
        _tracker.update();
        const eq::Matrix4f& headMatrix = _tracker.getMatrix();
        _setHeadMatrix( headMatrix );
    }

    // update database
    if( _animation.isValid( ))
    {
        const eq::Vector3f&  modelRotation = _animation.getModelRotation();
        const CameraAnimation::Step& curStep = _animation.getNextStep();

        _frameData.setModelRotation( modelRotation        );
        _frameData.setRotation(     curStep.rotation      );
        _frameData.setTranslation(  curStep.translation   );
    }
    else
    {
        if( _frameData.usePilotMode())
            _frameData.spinCamera( -0.001f * _spinX, -0.001f * _spinY );
        else
            _frameData.spinModel( -0.001f * _spinX, -0.001f * _spinY );

        _frameData.moveCamera( 0.0f, 0.0f, 0.001f*_advance );
    }

    [...]

    const uint32_t version = _frameData.commit();

    _redraw = false;
    return eq::Config::startFrame( version );
}
\end{lstlisting}}

@@

\begin{description}
\item[Buffer] The OpenGL read and draw buffer as well as color mask.
  These parameters are influenced by the current eye pass, eye
  separation and anaglyphic stereo settings.
\item[Viewport] The two-dimensional pixel viewport restricting the
  rendering area within the channel. For correct operations, both
  \textsf{glViewport} and \textsf{glScissor} have to be used. The pixel
  viewport is influenced by the destination channel's viewport
  definition and compound viewports set for sort-first/2D decompositions.
\item[Frustum] The same frustum parameters as defined by
  \textsf{glFrustum}. Typically the frustum used to set up the OpenGL
  projection matrix. The frustum is influenced by the destination
  channel's view definition, compound viewports, head matrix and the
  current eye pass. If the anti-aliasing idle mode is set, the frustum
  will be jittered by a vector. This part is described in \ref{sFSAA}.
\item[Head Transformation] A transformation matrix positioning the
  frustum. This is typically an identity matrix and is used for off-axis
  frusta in immersive rendering. It is normally used to set up the
  `view' part of the modelview matrix, before static light sources are
  defined.
\item[Range] A one-dimensional range with the interval [0..1]. This
  parameter is optional and should be used by the application to render
  only the appropriate subset of its data. It is influenced by the
  compound range attribute.
\end{description}

@@

{\footnotesize\begin{lstlisting}
    const Model* model = _getModel();
    if( model )
    {
        _drawModel( model );
    }
    else
    {
        glNormal3f( 0.f, -1.f, 0.f );
        glBegin( GL_TRIANGLE_STRIP );
        glVertex3f(  .25f, 0.f,  .25f );
        glVertex3f(  .25f, 0.f, -.25f );
        glVertex3f( -.25f, 0.f,  .25f );
        glVertex3f( -.25f, 0.f, -.25f );
        glEnd();
    }

    [...]
}
\end{lstlisting}}

@@

Like most applications, eqPly can use the default implementation of the
\textsf{frameReadback} and \textsf{frameAssemble} task methods.
\textsf{frameReadback} is overwritten for an optimization since eqPly
does not need the alpha channel on the destination view.
The output frames are flagged to ignore it, which allows the compressor to
drop 25\% of the data during image transfer:

@@

\textsf{frameAssemble} is overwritten for the use of Subpixel compound.
The whole process is described in \ref{sFSAA}.

@@

\subsection{\label{sFSAA}Full-scene and Software Anti-Aliasing}

The eqPly example contains support for software anti-aliasing to show 
how to combine it with the Equalizer Subpixel compound.

\subsubsection{FSAA or Supersampling algorithm}

The anti-aliasing algorithm uses here the supersampling technique which 
computes samples within a pixel using the jitter algorithm. The pixel is split
into several sub-pixels and a sample is taken randomly within the sub-pixel.

The sub-pixel area is a function of the current jitter step and the channel's
\textsf{subpixel} description. Jitter values for all channels should be evenly 
distributed over the total sub-pixel grid and we use therefore a precomputed
prime number table.

{\footnotesize\begin{lstlisting}
eq::Vector2i Channel::_getJitterStep() const
{
    uint32_t channelID  = getSubPixel().index;
    uint32_t idx = ( channelID * _totalSteps );
    idx += ( _jitterStep * primeNumberTable[ channelID ] ) \% _totalSteps;

    uint32_t sampleSize = _getSampleSize();
    const int dx = idx \% sampleSize;
    const int dy = idx / sampleSize;

    return eq::Vector2i( dx, dy );
}
\end{lstlisting}}

Next to that, sizes of a pixel and a sub-pixel on the near plane are computed.
A random position within the sub-pixel is set as a sample position and the 
vector values are correctly defined to finally be passed to the frustum.

{\footnotesize\begin{lstlisting}
eq::Vector2f Channel::_getJitter() const
{
    const eq::PixelViewport& pvp = getPixelViewport();
    float pvp_w = static_cast<float>( pvp.w );
    float pvp_h = static_cast<float>( pvp.h );
    float frustum_w = static_cast<float>(( getFrustum().get_width( )));
    float frustum_h = static_cast<float>(( getFrustum().get_height( )));

    float pixel_w = frustum_w / pvp_w;
    float pixel_h = frustum_h / pvp_h;

    uint32_t sampleSize = _getSampleSize();
    float subpixel_w = pixel_w / static_cast<float>( sampleSize );
    float subpixel_h = pixel_h / static_cast<float>( sampleSize );

    eq::Vector2i jitterStep = _getJitterStep();

    // Sample value randomly computed within the subpixel
    eq::base::RNG rng;
    float value_i = rng.get< float >() * subpixel_w
                    + static_cast<float>( jitterStep.x( )) * subpixel_w;

    float value_j = rng.get< float >() * subpixel_h
                    + static_cast<float>( jitterStep.y( )) * subpixel_h;

    return eq::Vector2f( value_i, value_j );
}
\end{lstlisting}}

\subsubsection{SWAA in eqPly}

This is the \textsf{FrameData} class which will give the information to the
application about being in idle mode or not. \textsf{Config} will update the 
idle mode information to the new \textsf{FrameData} version depending of the 
application state.

Since the application has a notion of idle/non-idle rendering, the 
\textsf{Channel} will be able to perform anti-aliasing when no user event is
detected and more frames are requested.

If the application is in idle mode, the \textsf{applyFrustum} operation will 
apply the computed jitter vector to the current frustum.

{\footnotesize\begin{lstlisting}
void Channel::applyFrustum() const
{
    const FrameData& frameData = _getFrameData();

    if( frameData.isIdle() && _jitterStep > 0 )
    {
        eq::Vector2f jitter = _getJitter();

        if( frameData.useOrtho( ))
        {
            eq::Frustumf ortho = getOrtho();
            ortho.apply_jitter( jitter );

            glOrtho( ortho.left(), ortho.right(),
                     ortho.bottom(), ortho.top(),
                     ortho.near_plane(), ortho.far_plane( ));
        }
        else
        {
            eq::Frustumf frustum = getFrustum();
            frustum.apply_jitter( jitter );

            glFrustum( frustum.left(), frustum.right(),
                       frustum.bottom(), frustum.top(),
                       frustum.near_plane(), frustum.far_plane( ));
        }
        return;
    }
    //else

    eq::Channel::applyFrustum();
}
\end{lstlisting}}

Then, \textsf{frameViewFinish} will get the frame buffer and accumulate 
it into an accumulation buffer. At each jitter step, the average result 
contained in the accumulation buffer will be displayed until all the steps
are done.
If the current pixel viewport is different from the one saved in 
\textsf{frameViewStart}, the accumulation buffer needs also to be resized 
and the idle anti-aliasing is reset.

{\footnotesize\begin{lstlisting}
void Channel::frameViewFinish( const uint32_t frameID )
{
    const FrameData& frameData = _getFrameData();

    const eq::PixelViewport& pvp = getPixelViewport();
    bool isResized = _accum->resize( pvp.w, pvp.h );

    if( isResized )
    {
        _accum->clear();
        _jitterStep = _totalSteps;
        _subpixelStep = 0;
        return;
    }
    //else

    if( frameData.isIdle() && _needsTransfer )
    {
        setupAssemblyState();

        _accum->accum();
        _accum->display();

        resetAssemblyState();
    }

    _drawOverlay();
    _drawHelp();

    if( frameData.useStatistics())
        EQ_GL_CALL( drawStatistics( ));
}
\end{lstlisting}}

At the end, \textsf{frameFinish} will decrement the jitter step and create a
new event to inform the \textsf{Config} of how much frames which remain to 
draw.

\subsubsection{SWAA in Equalizer}

As the software anti-aliasing decomposition should also work with the idle 
anti-aliasing in the application, the accumulation buffer is part of the 
\textsf{ObjectManager}. \textsf{frameAssemble} and \textsf{frameViewFinish} 
will then share the same accumulation buffer.

As for the application, the \textsf{eq::Channel::applyFrustum} will apply the
jitter vector if it exists. Then, if the function is not overwritten in the
application but if it still uses \textsf{subpixel} compounds, anti-aliasing 
will be correctly computed in Equalizer.

The main bit here is the assembly operation. \textsf{assembleFramesUnsorted} 
and \textsf{assembleFramesSorted} are similarly modified to handle the 
Subpixel decomposition.

For each Subpixel decomposition in input frames, all input frames are 
assembled with \textsf{subpixel} decomposition. The result is then accumulated
and displayed.

For example a 2-way subpixel compound configuration like 
\textsf{[0 2] [0 2] [1 2] [1 2]} will first assemble the both \textsf{[0 2]}
input frames and accumulate the result of the two assemblies. 
Similarly, the the both \textsf{[1 2]} frames are assembled and their result 
accumulated to the previous one.
In another hand, a 4-way subpixel compound configuration will do four 
assemblies and four accumulations.

{\footnotesize\begin{lstlisting}
void Compositor::assembleFramesUnsorted( const FrameVector& frames, 
                                         Channel* channel, Accum* accum )
{
    if( frames.empty( ))
        return;

    if( _isSubPixelDecomposition( frames ))
    {
        if( !accum )
        {
            accum = obtainAccum( channel );
            accum->clear();
        }

        FrameVector framesLeft = frames;
    	while( !framesLeft.empty( ))
    	{
            // Get the frames with the same subpixel compound
    	    FrameVector current = _extractOneSubPixel( framesLeft );
            assembleFramesUnsorted( current, channel, accum );
            accum->accum();
            accum->display();
        }
        return;
    }

    [...]

    // assemble frame
}
\end{lstlisting}}

@@

The viewport restricts the rendering to the area relative to the parent
compound. The range restricts the database range, relative to the
parent. The pixel setting selects the pixel decomposition kernel,
relative to the parent. The subpixel defines the jittering applied to
the frustum, relative to the parent. The eye pass select the mono, left
or right eye passes rendered by the compound. The zoom scales the parent
pixel viewport resolution. The DPlex period defines that $\frac{1}{period}$
frames are rendered, and the phase defines when in the period the
rendering starts. All these attributes are inherited by the children of
a compound. Viewport, range, pixel and period parameters are cumulative.

@@

    viewport [ viewport ]            # wrt parent compound, sort-first
    range    [ float float ]         # DB-range for sort-last
    pixel    [ int int ]             # pixel decomposition (step size)
    subpixel [ int int ]			 # subpixel decomposition (index size)
    eye      [ CYCLOP LEFT RIGHT ]   # monoscopic or stereo view
    zoom     [ float float ]         # up/downscale of parent pvp
    period   int                     # DPlex period, every nth frame rendered
    phase    int                     # DPlex phase, when to start rendering
